---
title: Plot Multiple Vector Layers
teaching: 40
exercises: 20
source: Rmd
---

```{r setup, echo=FALSE, results="hide"}
source("setup.R")
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

::::::::::::::::::::::::::::::::::::::: objectives

- Plot multiple vector layers in the same plot.
- Apply custom symbols to spatial objects in a plot.
- Create a multi-layered plot with raster and vector data.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I make different vecotr layers line up?
- How can I plot multiple forms of vector data together?

::::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::::::::  prereq

## Things You'll Need To Complete This Episode

See the [lesson homepage](.) for detailed information about the software, data,
and other prerequisites you will need to work through the examples in this
episode.


::::::::::::::::::::::::::::::::::::::::::::::::::

This episode builds upon
[the previous episode](07-vector-shapefile-attributes-in-r/)
to work with vector layers in R and explore how to plot multiple
vector layers. It also covers how to plot raster and vector data together on the
same plot.

## Load the Data

To work with vector data in R, we can use the `sf` library. We will also be using `ggplot2` and some `dplyr`. Let's start by loading them.

```{r load-libraries, echo=TRUE, results="hide", message=FALSE}
library(ggplot2)
library(dplyr)
library(sf)
```

We are going to plot the 2022 Casco Bay Seagrass Beds, but, include some coastline for context. We will also add layers of information along the way. To start with, let's load up the seagrass beds, the AOI for Casco, and a shapefile of counties in Maine.

```{r load-data, echo=TRUE, results="hide", warning=FALSE}
aoi_boundary_casco <- st_read(
  "data/maine_gov_maps/casco_aoi/casco_bay_aoi.shp")

seagrass_casco_2022 <- st_read(
  "data/maine_gov_seagrass/MaineDEP_Casco_Bay_Seagrass_2022/MaineDEP_Casco_Bay_Seagrass_2022.shp")

maine_borders <- st_read(
  "data/maine_gov_maps/Maine_State_Boundary_Polygon_Feature/Maine_State_Boundary_Polygon_Feature.shp")
```

## Making Sure Layers Match in Extent

One of the beautiful things about `ggplot2` is that we're going to be able to just add layers together to make a nice plot. Our goal here is to plot seagrass beds with a coastline bordering them, so we can see how they line up around islands. However, we have a small problem. Let's compare the extent of our objects using `st_bbox()`.

```{r show-ext}
st_bbox(aoi_boundary_casco)
st_bbox(seagrass_casco_2022)
st_bbox(maine_borders)
```

While the first two are close - indeed, the AOI was made from the seagrass polygons extent - the state of Maine is huge relative to Casco Bay. We need to **crop** that shapefile down to just the area we want to plot. Fortunately, `sf` features a function called `st_crop()` which will crop a large vector file down to the size of the extent of a smaller vector file. So if we just want the Casco Coastline, we can crop `maine_borders` down to the size of the `aoi_boundary_casco` vector file.

```{r crop_bad, error=TRUE}
casco_coastline <- st_crop(maine_borders, aoi_boundary_casco)
```

This does not always go well with data sets from the wild. Without getting too deep into it, vector files (particularly polygons) can have a variety of issues in them which make them invalid. To fix this, we need to make them valid before cropping with `st_make_valid()`.  Odd, but, this is incredibly common.

```{r crop, warning = FALSE}
casco_coastline <- st_crop(maine_borders |> st_make_valid(), 
                           aoi_boundary_casco)

ggplot() +
  geom_sf(data = casco_coastline) +
  coord_sf() +
  theme_void()
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Casco Bay Region Roads

Load up the roads of Maine and crop them to the Casco Bay region.

:::::::::::::::  solution

```{r load_crop_roads, warning = FALSE}
roads_maine <- st_read("data/maine_gov_maps/MaineDOT_Public_Roads/MaineDOT_Public_Roads.shp")

roads_casco <- st_crop(roads_maine, aoi_boundary_casco)

ggplot() +
  geom_sf(data = roads_casco)
```

:::::::::::::::::::::::::::::::::::::::  
:::::::::::::::::::::::::::::::::::::::  


Lovely! Now let's start bringing seagrass in.

## Plotting Multiple Vector Layers

In the [previous episode](07-vector-shapefile-attributes-in-r/), we learned how
to plot information from a single vector layer and do some plot customization
including adding a custom legend. However, what if we want to create a more
complex plot with many vector layers and unique symbols that need to be
represented clearly in a legend?

Now, let's create a plot that combines our seagrass beds 
(`seagrass_casco_2022`), the coastline (`casco_coastline`) and roads 
(`roads_maine`) spatial objects. We will need to build a custom legend as well.

To begin, we will create a plot with the coastline as the base layer with a tan 
fill and a black outline. We will 
add on top roads, but with alpha = 0.1 so they are faint. 
Last, we will layer seagrass beds on top with color and fill as percent cover 
and a linewidth of 1.3. We will save it as an object so we can make small 
changes from here on out.

```{r plot-many-shapefiles}
seagrass_map <- ggplot() +
  geom_sf(data = casco_coastline, fill = "tan", color = "black") +
  geom_sf(data = roads_casco, alpha = 0.1) +
  geom_sf(data = seagrass_casco_2022, 
          mapping = aes(color = Cover_Pct,
                        fill = Cover_Pct),
          linewidth = 1.3) +
  scale_color_brewer(palette = "Greens") +
  scale_fill_brewer(palette = "Greens") 

seagrass_map
```

This looks OK, but, let's dial things up a bit. Let's

1. Eliminate the gap around the plot with an argument `expand = FALSE` to `coord_sf()`.
2. Give it a cleaner theme with `theme_classic()`
3. Give the map a name and the fill/color legend a name.
4. Use `theme()` to make it more ocean-y with a 

We can start by making the plot cleaner and eliminating the gap (1 & 2)

```{r plot-custom-shape}
seagrass_map <- seagrass_map +
  coord_sf(expand = FALSE) +
  theme_classic()

seagrass_map
```

Now lets adjust the legend titles by using `labs()`.

```{r create-custom-legend}
seagrass_map <- seagrass_map + 
  labs(color = "% Cover", fill = "% Cover", 
       title = "Seagrass in Casco Bay in 2022")

seagrass_map
```

Last, we can use `theme()` to add a light blue background and rotate the
X axis. We will save this as an object, so we don't need to type more as we make further modifications.


```{r better_theme}
seagrass_map <- seagrass_map  +
  theme(panel.background = element_rect(fill = "lightblue"),
        axis.text.x = element_text(angle = 45, hjust = 1))

seagrass_map
```

## Zooming In

This map looks great, but, there's a lot of data here. While it can provide a great big-picture overview, it's hard to see more small-scale features. There are two solutions to this. The first is plotting the map zoomed in to an area of interest. You can get your AOI either by eyeballing the corner coordinates (x and y 
min and max) from the plot itself, or use something like 
[https://maps.google.com](https://maps.google.com) and right click to get point
coordinates.

Using the later method, here, for example, is the area around Mackworth Island.

```{r coord_sf_zoom}
seagrass_map +
  coord_sf(xlim = c(-70.24768051612979, -70.19432151530208),
           ylim = c(43.66353031177649, 43.714022044471584),
           expand = FALSE)
```

The second answer is an interactive map with leaflet. This is slightly more challenging to implement, but very similar to what we did with rasters before. We will start by making a color palette for the map. We can use `colorFactor()` as we have discrete classes here.

```{r leaflet}
library(leaflet)

seagrass_pal <- colorFactor(palette = "Greens",
                         domain = seagrass_casco_2022$Cover_Pct)
```

From here, we can build a leaflet map. Let's use `addProviderTiles("Stadia.AlidadeSmooth")` for a very neutral background. We add
`sf` polygons using `addPolygons()`. For this, we need to think about the what
is creating polygon borders and the fill. We will set `stroke = FALSE` so we don't have to worry about the border (too many more argument) and instead use our 
palette - which is now a function to be evaluated with `~` and set a 
`fillOpacity` to 1, for fully opaque.

```{r leaflet_map}
leaflet() |>
 addProviderTiles("Stadia.AlidadeSmooth") |>  
  addPolygons(data = seagrass_casco_2022, 
              fillColor = ~seagrass_pal(Cover_Pct),
              fillOpacity = 1,
              stroke = FALSE)
```

Note, you can recreate what you did with the `ggplot2` above just using 
the `sf objects`. We'll need to invoke one trick from the `leaflet.extras` 
package to get that blue background.

```{r leaflet_custom}
library(leaflet.extras)

leaflet() |>
  addPolygons(data = casco_coastline, 
              fillColor = "tan",
              fillOpacity = 1,
              stroke = FALSE) |>
  addPolylines(data = roads_casco,
               color = "black",
               opacity = 0.1,
               weight = 1) |>
  addPolygons(data = seagrass_casco_2022, 
              fillColor = ~seagrass_pal(Cover_Pct),
              fillOpacity = 1,
              stroke = FALSE) |>
  addPolygons(data = aoi_boundary_casco,
              color = "black",
              fillOpacity = 0) |>
  setMapWidgetStyle(list(background= "lightblue")) |>
  addLegend(pal = seagrass_pal, 
            values = seagrass_casco_2022$Cover_Pct,
            title = "% Cover Seagrass")
```


:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge: Let's Part Like it's 1997... er, 1993-1994

1. The earliest seagrass datat available from Maine's Geolibrary is their [1997 State-Wide Data](https://maine.hub.arcgis.com/datasets/maine::mainedmr-eelgrass-1997/about) aggregated from photography from 1992 - 1997. Load up the ESRI `shapefile` from `data/maine_gov_seagrass/MaineDMR_-_Eelgrass-1997-shp` and do a quick visualization showing where was surveyed when.

2. Crop it down to Casco Bay and make a plot showing the area around Mackworth Island. Make sure your title reflects what years are being represented. While there is a cover classification (from 0-5) based on percent cover, the metadata doesn't have the lineup. Still, qualitatively, how does this compare to 2022? Use whatever technique you'd like to visualize this small area. Note, `COVER` is continuous, so to make it into distinct classes you will need `as.character(COVER)` in your code.

:::::::::::::::  solution

## Answers


1. First we need to load in the data. Let' see how many years are here.

```{r}
seagrass_1997 <- st_read("data/maine_gov_seagrass/MaineDMR_-_Eelgrass-1997-shp/MaineDMR_-_Eelgrass.shp")

# what are the names
names(seagrass_1997)

# what are the unique years
unique(seagrass_1997$Year97)
```

Then, we can visualize. 

```{r}
ggplot() +
  geom_sf(data = seagrass_1997,
          aes(color = Year97, fill = Year97)) +
  scale_color_viridis_c() +
  scale_fill_viridis_c() 
```

2. For the second piece, first, we need to crop to Casco Bay and check what years this data consists of.
```{r}
seagrass_casco_1997 <- st_crop(seagrass_1997 |> st_make_valid(),
                              aoi_boundary_casco)

unique(seagrass_casco_1997$Year97)
```

With the cropped polygons, we can make a map akin to what we did above for 2022. We can even re-use our code, but with a few small modifications for a different data set and a different cover type.

```{r}
ggplot() +
  geom_sf(data = casco_coastline, fill = "tan", color = "black") +
  geom_sf(data = roads_casco, alpha = 0.1) +
  geom_sf(data = seagrass_casco_1997, 
          mapping = aes(color = as.character(COVER),
                        fill = as.character(COVER)),
          linewidth = 1.3) +
  scale_color_brewer(palette = "Greens") +
  scale_fill_brewer(palette = "Greens") + 
  theme_classic() +
  labs(color = "Cover Class", fill = "Cover Class", 
       title = "Seagrass in Casco Bay in 1993 and 1994") +
  coord_sf(xlim = c(-70.24768051612979, -70.19432151530208),
           ylim = c(43.66353031177649, 43.714022044471584),
           expand = FALSE) +
  theme(panel.background = element_rect(fill = "lightblue"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

While we cannot know how the beds compare exactly, given that we do not know how the Cover Class maps to % Cover, it appears that many of the beds in 1993 and 1994 are at the highest class of cover, while in 2022 they are at the high-middle.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::::::: keypoints

- Use the `+` operator to add multiple layers to a ggplot.
- Use `st_crop()` to put spatially subset a vector data set.
- Multi-layered plots can combine multiple vector data sets.
- Use `leaflet` or zooming in `ggplot2` to see small spatial features.

::::::::::::::::::::::::::::::::::::::::::::::::::


